Adopting agile software development in the large organizations: A systematic literature review

### 2014-08-11 journal/content.tex
Fixed citation tags

* There is interest in large scale [ref XP13 workshop] \cite{Dingsoyr2013a}

@@

Agile software development is a collection of methods developed as an alternative
to so called traditional development methods. From the point of view of agile
methods traditional methods strive to minimize change during the development
process and attempt to fully define requirements before development. However,
adaptability during the development life cycle and the ability to respond to
conditions unforeseen at start are critical factors for the success of a
software development project. Embracing change during development, people
centricity are, and pursuing high quality from the start are fundamental themes
in agile development.
\cite{Highsmith2001,Cockburn2001}

Agile methods have been both criticized and advocated, and research has shown
that accommodating change may be a factor in both success and failure
\cite{Boehm2002}. It has been shown that agile methods have improved
satisfaction of both customers and developers, but on the other hand there is
evidence that agile methods may not suite well for large undertakings
\cite{Dyba2009}. As a solution \citet{Boehm2002} has suggested that each
organization should find its own balance of agile and plan driven methods.

Two of the most popular agile methods at the time being are Extreme Programming
(XP) and Scrum \cite{Hamed2013}. Scrum is a method which focuses on the project
management viewpoint of agile development \cite{Schwaber2002}. The Scrum method
focuses on time-boxing, continuous tracking of project progress, and customer
centricity. A prominent feature of Scrum is dividing work into sprints, which
are time boxes of 1 to 4 weeks, aimed for producing a finished increment of
working software \cite{Schwaber2002}. The XP development method is a collection
of practices for enabling efficient incremental development \cite{Beck1999}.
Some of the key practices of XP are continuous feedback on development, test
driven development, close customer involvement, pair programming, and continuous
integration (CI) \cite{Beck1999}.

A methodology relating to and complementing agile development is lean software
development. Similarly to Scrum and XP, lean software development aims to
concentrate focus on activities that provide the most value to the customer.
Core principles of lean software development are optimizing the whole, removal
of waste, building quality in, and continuous learning. \cite{Poppendieck2012}

@@

Introducing agile methods in large organizations is more difficult than it is in
small organizations \cite{Dyba2008}. The difficulty is partly related to size
bringing higher organizational inertia which slows down organizational change
\cite{Livermore2008}. Agile development is not founded on the use of individual
tools or practices, but rather on a holistic way of thinking. Adopting agile
often requires change of the entire organizational culture \cite{Misra2010}.

One significant difference between small and large scale adoptions is that large
organizations have more dependencies between projects and teams. This increases
the need for formal documentation and thus reduces agility \cite{Lindvall2004}.
In addition to inter-team coordination development teams must interact with
other organizational units, which are often non-agile in nature. For instance,
HR may demand individuals to have strictly specified roles in projects
\cite{Boehm2005}, or a change control board may inhibit the use of continuous
integration or refactoring \cite{Lindvall2004}. All units affected by the agile
transformation need to be informed and consulted, and the agile process must be
adjusted according to their needs \cite{Lindvall2004, Cohn2003, Boehm2005}.

Agile methods will also affect management and business related functions. A key
challenge is that management must move away from life-cycle models and towards
iterative and feature centric models \cite{Nerur2005}. The focus must be shifted
from a large scale scope to shorter term project planning \cite{Misra2010}, as
agile methods emphasize that planning is only meaningful for the near future
\cite{Cohn2003}. However, the lack of planning can be a concern as business and
customer relationships often build on long term roadmapping. Enabling operation
with shorter term planning requires educating stakeholders and reviewing
contracting practices \cite{Boehm2005}.

@@

The definition of large scale was recently discussed in a workshop at XP2013
\cite{Dingsoyr2013a}. Proceeding the workshop, Dingsøyr et al.
compiled a description of the definition of large scale \cite{Dingsoyr2013b}.
The most common view in the XP2013 workshop was that the involvement of many
people is the main factor in large scale \cite{Dingsoyr2013b}.

Dingsøyr et al. conducted a brief literature search to identify what definitions
on large scale are used \cite{Dingsoyr2013b}. The results showed that large
scale had been regarded in terms of size in persons or teams, project budget,
code base size, and project duration. As examples from the results, Paasivaara
et al. \cite{Paasivaara2008} referred to an organization of 40 people and 7
teams. Berger and Beynon-Davies \cite{Berger2009} present the project cost of
over £10 million as a factor of large scale, but also highlight a team size of
over 50 people. A product with a code base size of over 5 million lines of code
was considered to represent a large scale development effort by Petersen and
Wohlin \cite{Petersen2010}. A project time of 2 years was considered as a factor
of large scale by Bjarnson et al. \cite{Bjarnason2011}, as well as a project
scope of 60-80 features.

As a part of our research, we identified further definitions of large scale. All
of these definitions referred to the number of people involved.
In early work on agile Fowler \cite{Fowler2000} regards the Crystal methodology
being suitable for up to 50 people, thus setting a boundary beyond which agile
methods would be expected to need special considerations. Also Williams and
Cockburn \cite{Williams2003} present 50 people as a limit which practitioners
and researchers has seen as the largest organization suitable for agile.
Elshamy and Elssamadisy \cite{Elshamy2006} considered a development project
large if it had a staff between 50 and 100 people, including all project
personnel. Finally, the largest numbers were presented by Moore and Spens
\cite{Moore2008} who define large scale in the scope of their research as 300
people across 3 sites.

@@

The research was conducted as an application of Kitchenham's
\cite{Kitchenham2007} guidelines for systematic literature review. Selection of
primary studies was done using first a keyword based database search to identify
potentially relevant sources, and then manually filtering the search result. The
filtering process was executed independently by two researchers (Dikert,
Paasivaara). The selected primary studies were coded for qualitative data
extraction by one researcher (Dikert). The results of the study were finally
elicited by aggregating and analyzing the coding of the primary documents.
Figure \ref{fig:researchprocess} presents an outline of the research process.
The entire process was audited and mentored by a third researcher (Lassenius).

@@

Examples on topics excluded by the facet of \emph{agile software development}
are agile manufacturing and applying Scrum practices in management boards, as
these do not relate to software engineering. In addition, we required that the
organizational transformation was aimed to introduce agile methods, which
excluded other development methods than agile \cite{Sagesser2013}, and use of
agile methods in other contexts than software development \cite{Hodgkins2007}.

The facet of \emph{organizational transformation} was interpreted so that the
primary study must present insights on the process of transformation. This
was the most important facet, as the pivot of this research is to examine
how the transformations happen. Examples on excluded topics are comparing the
original and agile development methods \cite{Petersen2010}, discussing use of
agile in a large organization but not describing how the new methods were
introduced \cite{Mishra2011}, and merely presenting agile tools in large-scale
use \cite{Kim2012}. These sources were excluded as they did not provide insights
on organizational transformation.
% Also [Fitzgerald 2005] on just presenting use of agile methods, no introducing

Transformation and ``scaling up'' of agile practices in use are very closely
related concepts, and in some cases they are one and the same. For instance, if
transformation begins with a pilot and spreads gradually through the
organization, the process can very well be seen as a ``scaling up'' journey.
However, we excluded cases where an initially small organization was scaled up
\cite{Maranzato2012}, and discussions focusing on processes or tools without
describing organizational change \cite{Lyon2008}.
% Also [Greening 2010] on process centered viewpoint without transformation

The facet of \emph{large scale} was interpreted as presented in section
\ref{sec:largescale}, having 50 development personnel or 6 teams as as the limit
for large scale. In some cases the source presented very vague indicators on
size. For instance, we deemed the case presented by Cloke \cite{Cloke2007} to be
included as there was indications of large-scale considerations, although the
size remained unclear. The case presented by Miller and Haddad \cite{Miller2012}
was deemed to be excluded as there was no indications whatsoever on size. A
borderline exclude was the case presented by Tudor \cite{Tudor2006} as there was
no indication of issues indicating large size, although there were several
development teams.
To complicate matters, some papers talked about ``the team'' in singular when
referring to the organization \cite{Hodgkins2007}, making it clearly nontrivial
to judge whether an organization was large based on the choice of words of the
author.

Further examples on exclusion by the \emph{large scale} facet were cases with
large organizations but only a single team adopting agile \cite{Fulgham2011}.
We considered cases of single teams (although in large organizations) unrelated
to this research as our focus is on transformation of the entire organization.
Also piloting cases that resulted only in single teams using agile were excluded
\cite{Scott2008}. Finally, cases where the organization was growing to large
scale, but did not meet the size criteria at the start of the transformation,
were excluded.

@@

In several cases there were more than one primary study presenting the
transformation of the same organization. Duplicate descriptions of the same
organization focused typically on different aspects. For example, one paper
would highlight the viewpoint of developers \cite{Fry2007}, and another would
consider the transformation from user experience designers' point of view
\cite{Federoff2009}.

@@

As presented in the previous section, we allowed several descriptions of the
same case to be included. Instead of using the most complete paper as suggested
by Kitchenham \cite{Kitchenham2007}, we combined the results presented in each
paper and considered the case as a single unit. Keeping in mind the potential
bias caused by duplicate publications, including all papers enabled us to have a
more detailed view of the case organization.

@@

The primary studies were coded using the Atlas TI qualitative data analysis software.
An integrated deductive and inductive approach was chosen for coding the primary
studies \cite{Cruzes2011a}. The coding was designed to have a contextual part
and a findings part as also suggested by Cruzes and Dybå \cite{Cruzes2011a}. A
list of codes was established for contextual information, as presented in Table
\ref{table:contextualcodes}. Codes related to the research questions were chosen
to be created by an inductive process. The reason for inducting codes rather
than using an a priori approach was to avoid the researchers' previous
assumptions of the research area to affect the choice of codes

@@

To give direction for code induction seven different code families were
established on beforehand, as presented in Table \ref{table:codefamilies}. A
description defining scope and examples for codes were defined for each family.
The example codes would not necessarily be used as actual codes, and would not
represent the entire or final scope of the families.

@@

Coding was done by only one researcher due to resource constraints and volume of
source material. As recommended by Kitchenham \cite{Kitchenham2007} we conducted
an experiment with the goal to assess the consistency of coding that two
independent researchers may achieve.
In the experiment Dikert and Paasivaara coded independently a sample of five
primary studies. The sample was hand picked as a representative subset of the
source material. The coding was done relying solely on the guidelines presented
in the previous chapter, and the researchers were careful not to discuss the
choice of codes or interpretations relating to codes before the experiment was
completed. Finally the independent codings of the two researchers were analyzed
with two metrics: by the number of matching themes found, and by the amount of
overlap in quotations similar codes have.
% See Kitcenham, chapter 6.4

@@

The old model was described in most cases briefly and in a neutral sentiment.
Waterfall development in general was referred to as ``traditional, plan-based
methods'' [P25], ``traditional phase-based approach'' [P36], or ``old fashioned
process driven waterfall'' [P23]. Korhonen [P25] referred to the model presented
by Royce \cite{Royce1970} when describing the initial state.

@@

Transformation can be facilitated by creating mechanisms that force change.
Spayd [P49] describes that senior management pushed hard on a mandate to have
deliveries every 90 days. The mandate made change necessary, and while the
strong pressure did not promote agile practices in every case, the drive for
change was perceived good in general [P49].

>>>>

* There is interest in large scale [ref XP13 workshop] \citep{Dingsoyr2013a}

@@

Agile software development is a collection of methods developed as an alternative
to so called traditional development methods. From the point of view of agile
methods traditional methods strive to minimize change during the development
process and attempt to fully define requirements before development. However,
adaptability during the development life cycle and the ability to respond to
conditions unforeseen at start are critical factors for the success of a
software development project. Embracing change during development, people
centricity are, and pursuing high quality from the start are fundamental themes
in agile development.
\citep{Highsmith2001,Cockburn2001}

Agile methods have been both criticized and advocated, and research has shown
that accommodating change may be a factor in both success and failure
\citep{Boehm2002}. It has been shown that agile methods have improved
satisfaction of both customers and developers, but on the other hand there is
evidence that agile methods may not suite well for large undertakings
\citep{Dyba2009}. As a solution \citet{Boehm2002} has suggested that each
organization should find its own balance of agile and plan driven methods.

Two of the most popular agile methods at the time being are Extreme Programming
(XP) and Scrum \citep{Hamed2013}. Scrum is a method which focuses on the project
management viewpoint of agile development \citep{Schwaber2002}. The Scrum method
focuses on time-boxing, continuous tracking of project progress, and customer
centricity. A prominent feature of Scrum is dividing work into sprints, which
are time boxes of 1 to 4 weeks, aimed for producing a finished increment of
working software \citep{Schwaber2002}. The XP development method is a collection
of practices for enabling efficient incremental development \citep{Beck1999}.
Some of the key practices of XP are continuous feedback on development, test
driven development, close customer involvement, pair programming, and continuous
integration (CI) \citep{Beck1999}.

A methodology relating to and complementing agile development is lean software
development. Similarly to Scrum and XP, lean software development aims to
concentrate focus on activities that provide the most value to the customer.
Core principles of lean software development are optimizing the whole, removal
of waste, building quality in, and continuous learning. \citep{Poppendieck2012}

@@

Introducing agile methods in large organizations is more difficult than it is in
small organizations \citep{Dyba2008}. The difficulty is partly related to size
bringing higher organizational inertia which slows down organizational change
\citep{Livermore2008}. Agile development is not founded on the use of individual
tools or practices, but rather on a holistic way of thinking. Adopting agile
often requires change of the entire organizational culture \citep{Misra2010}.

One significant difference between small and large scale adoptions is that large
organizations have more dependencies between projects and teams. This increases
the need for formal documentation and thus reduces agility \citep{Lindvall2004}.
In addition to inter-team coordination development teams must interact with
other organizational units, which are often non-agile in nature. For instance,
HR may demand individuals to have strictly specified roles in projects
\citep{Boehm2005}, or a change control board may inhibit the use of continuous
integration or refactoring \citep{Lindvall2004}. All units affected by the agile
transformation need to be informed and consulted, and the agile process must be
adjusted according to their needs \citep{Lindvall2004, Cohn2003, Boehm2005}.

Agile methods will also affect management and business related functions. A key
challenge is that management must move away from life-cycle models and towards
iterative and feature centric models \citep{Nerur2005}. The focus must be shifted
from a large scale scope to shorter term project planning \citep{Misra2010}, as
agile methods emphasize that planning is only meaningful for the near future
\citep{Cohn2003}. However, the lack of planning can be a concern as business and
customer relationships often build on long term roadmapping. Enabling operation
with shorter term planning requires educating stakeholders and reviewing
contracting practices \citep{Boehm2005}.

@@

The definition of large scale was recently discussed in a workshop at XP2013
\citep{Dingsoyr2013a}. Proceeding the workshop, Dingsøyr et al.
compiled a description of the definition of large scale \citep{Dingsoyr2013b}.
The most common view in the XP2013 workshop was that the involvement of many
people is the main factor in large scale \citep{Dingsoyr2013b}.

Dingsøyr et al. conducted a brief literature search to identify what definitions
on large scale are used \citep{Dingsoyr2013b}. The results showed that large
scale had been regarded in terms of size in persons or teams, project budget,
code base size, and project duration. As examples from the results,
\citet{Paasivaara2008} referred to an organization of 40 people and 7 teams.
\citet{Berger2009} present the project cost of over £10 million as a factor of
large scale, but also highlight a team size of over 50 people. A product with a
code base size of over 5 million lines of code was considered to represent a
large scale development effort by \citet{Petersen2010}. A project time of 2
years was considered as a factor of large scale by \citet{Bjarnason2011}, as
well as a project scope of 60-80 features.

As a part of our research, we identified further definitions of large scale. All
of these definitions referred to the number of people involved.
In early work on agile \citet{Fowler2000} regards the Crystal methodology being
suitable for up to 50 people, thus setting a boundary beyond which agile methods
would be expected to need special considerations. Also \citet{Williams2003}
present 50 people as a limit which practitioners and researchers has seen as the
largest organization suitable for agile.
\citet{Elshamy2006} considered a development project large if it had a staff
between 50 and 100 people, including all project personnel. Finally, the largest
numbers were presented by \citet{Moore2008} who define large scale in the scope
of their research as 300 people across 3 sites.

@@

The research was conducted as an application of the guidelines for systematic
literature review presented by \citet{Kitchenham2007}. Selection of primary
studies was done using first a keyword based database search to identify
potentially relevant sources, and then manually filtering the search result. The
filtering process was executed independently by two researchers (Dikert,
Paasivaara). The selected primary studies were coded for qualitative data
extraction by one researcher (Dikert). The results of the study were finally
elicited by aggregating and analyzing the coding of the primary documents.
Figure \ref{fig:researchprocess} presents an outline of the research process.
The entire process was audited and mentored by a third researcher (Lassenius).

@@

Examples on topics excluded by the facet of \emph{agile software development}
are agile manufacturing and applying Scrum practices in management boards, as
these do not relate to software engineering. In addition, we required that the
organizational transformation was aimed to introduce agile methods, which
excluded other development methods than agile \citep{Sagesser2013}, and use of
agile methods in other contexts than software development \citep{Hodgkins2007}.

The facet of \emph{organizational transformation} was interpreted so that the
primary study must present insights on the process of transformation. This
was the most important facet, as the pivot of this research is to examine
how the transformations happen. Examples on excluded topics are comparing the
original and agile development methods \citep{Petersen2010}, discussing use of
agile in a large organization but not describing how the new methods were
introduced \citep{Mishra2011}, and merely presenting agile tools in large-scale
use \citep{Kim2012}. These sources were excluded as they did not provide insights
on organizational transformation.
% Also [Fitzgerald 2005] on just presenting use of agile methods, no introducing

Transformation and ``scaling up'' of agile practices in use are very closely
related concepts, and in some cases they are one and the same. For instance, if
transformation begins with a pilot and spreads gradually through the
organization, the process can very well be seen as a ``scaling up'' journey.
However, we excluded cases where an initially small organization was scaled up
\citep{Maranzato2012}, and discussions focusing on processes or tools without
describing organizational change \citep{Lyon2008}.
% Also [Greening 2010] on process centered viewpoint without transformation

The facet of \emph{large scale} was interpreted as presented in section
\ref{sec:largescale}, having 50 development personnel or 6 teams as as the limit
for large scale. In some cases the source presented very vague indicators on
size. For instance, we deemed the case presented by \citet{Cloke2007} to be
included as there was indications of large-scale considerations, although the
size remained unclear. The case presented by \citet{Miller2012} was deemed to be
excluded as there was no indications whatsoever on size. A borderline exclude
was the case presented by \citet{Tudor2006} as there was no indication of issues
indicating large size, although there were several development teams.
To complicate matters, some papers talked about ``the team'' in singular when
referring to the organization \citep{Hodgkins2007}, making it clearly nontrivial
to judge whether an organization was large based on the choice of words of the
author.

Further examples on exclusion by the \emph{large scale} facet were cases with
large organizations but only a single team adopting agile \citep{Fulgham2011}.
We considered cases of single teams (although in large organizations) unrelated
to this research as our focus is on transformation of the entire organization.
Also piloting cases that resulted only in single teams using agile were excluded
\citep{Scott2008}. Finally, cases where the organization was growing to large
scale, but did not meet the size criteria at the start of the transformation,
were excluded.

@@

In several cases there were more than one primary study presenting the
transformation of the same organization. Duplicate descriptions of the same
organization focused typically on different aspects. For example, one paper
would highlight the viewpoint of developers \citep{Fry2007}, and another would
consider the transformation from user experience designers' point of view
\citep{Federoff2009}.

@@

As presented in the previous section, we allowed several descriptions of the
same case to be included. Instead of using the most complete paper as suggested
by \citet{Kitchenham2007}, we combined the results presented in each
paper and considered the case as a single unit. Keeping in mind the potential
bias caused by duplicate publications, including all papers enabled us to have a
more detailed view of the case organization.

@@

The primary studies were coded using the Atlas TI qualitative data analysis
software. An integrated deductive and inductive approach was chosen for coding
the primary studies \citep{Cruzes2011a}. The coding was designed to have a
contextual part and a findings part as also \citet{Cruzes2011a} suggests. A list
of codes was established for contextual information, as presented in Table
\ref{table:contextualcodes}. Codes related to the research questions were chosen
to be created by an inductive process. The reason for inducting codes rather
than using an a priori approach was to avoid the researchers' previous
assumptions of the research area to affect the choice of codes.

@@

To give direction for code induction, seven different code families were
established on beforehand, as presented in Table \ref{table:codefamilies}. A
description defining scope and examples for codes were defined for each family.
The example codes would not necessarily be used as actual codes, and would not
represent the entire or final scope of the families.

@@

Coding was done by only one researcher due to resource constraints and volume of
source material. As recommended by \citet{Kitchenham2007} we conducted
an experiment with the goal to assess the consistency of coding that two
independent researchers may achieve.
In the experiment Dikert and Paasivaara coded independently a sample of five
primary studies. The sample was hand picked as a representative subset of the
source material. The coding was done relying solely on the guidelines presented
in the previous chapter, and the researchers were careful not to discuss the
choice of codes or interpretations relating to codes before the experiment was
completed. Finally the independent codings of the two researchers were analyzed
with two metrics: by the number of matching themes found, and by the amount of
overlap in quotations similar codes have.
% See Kitcenham, chapter 6.4

@@

The old model was described in most cases briefly and in a neutral sentiment.
Waterfall development in general was referred to as ``traditional, plan-based
methods'' [P25], ``traditional phase-based approach'' [P36], or ``old fashioned
process driven waterfall'' [P23]. Korhonen [P25] referred to the model presented
by \citet{Royce1970} when describing the initial state.

@@

Transformation can be facilitated by setting up mechanisms that force change.
Spayd [P49] describes that senior management pushed hard on a mandate to have
deliveries every 90 days. The mandate made change necessary, and while the
strong pressure did not promote agile practices in every case, the drive for
change was perceived good in general [P49].
